---
output:
  html_document:
    fig_height: 8
    fig_width: 15
---
# Reproducible Research - Exploring & Analyzing NOAA Storm Database
### Author: Himanshu Rawat
### Date: 20th June, 2015

******

- [Synopsis](#synopsis)
- [Data Processing](#data-processing)
    - [FATALITIES Processing](#fatalities-processing)
    - [INJURIES Processing](#injuries-processing)
    - [ECONOMIC Processing](#economic-processing)
- [Results](#results)

***********

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(clean=FALSE)
```

## Synopsis

The basic goal of this assignment is to explore NOAA Storm Database for the severity of the weather events. We will also answer below questions:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

[Storm Data](http://d396qusza40orc.cloudfront.net/repDF%2FDF%2FStormDF.csv.bz2)

## Data Processing

I am assuminng data has been downloaded and extracted via above link and configured correct current working directory.  
Load neccessary packages.
```{r}
library(package = "plyr", warn.conflicts = FALSE, quietly = TRUE)
library(package = "dplyr", warn.conflicts = FALSE, quietly = TRUE)
library(package = "ggplot2", warn.conflicts = FALSE, quietly = TRUE)
```
Simple read of CSV file as we need to see all the variables. I will limit to 10 rows.
```{r}
storm.data <- read.csv(file ="repdata_data_StormData.csv", nrow = 10)
str(storm.data)
```
Our raw storm data consists of <span style="color:red;">`37`</span> columns but we are interested in <span style="color:red;">`8`</span> columns i.e.

Variale name    |   Description
----------      |   ------------
STATE           |   Two letter code
EVTYPE          |   Weather event source type (tornado,avalanche,ice-storm etc.)    
FATALITIES      |   Count of weather event related fatalities
INJURIES        |   Count of weather event related injuries
PROPDMG         |   Property damage (scaled value)
PROPDMGEXP      |   Unit of measure for the PROPDMG
CROPDMG         |   Crop damage (scaled value)
CROPDMGEXP      |   Unit of measure for CROPDMG

Lets read only above columns
```{r}
columns.to.read <- c(rep("NULL",6), rep("character",2) , 
                        rep("NULL",14),rep("numeric",3),"character","numeric","character", rep("NULL",9))

storm.data <- read.csv(file ="repdata_data_StormData.csv", colClasses = columns.to.read , nrow = 1232705)
```
### Pre process the data for simplifications
```{r}
# for every state, get the total counts of each event type
new.storm <- storm.data %>% group_by(STATE,EVTYPE) %>% summarise(fatal.count = sum(FATALITIES), injur.count = sum(INJURIES), total.count = sum(FATALITIES + INJURIES))
```
Harmful events across the states w.r.t population health can be divided into two parts

1. FATALITIES
2. INJURIES

Economic consequences will use below columns

1. PROPDMG
2. PROPDMGEXP
3. CROPDMG
4. CROPDMGEXP

### <span style="color:red;">`FATALITIES`</span> Processing
```{r}
# get the event have max count and remove duplicates
fatal <- new.storm %>% top_n(n = 1, wt = fatal.count) %>% distinct()
# summarise the total count of each event. This will give us unqiue fatalities.
fatal.new <- fatal %>% group_by(EVTYPE) %>% summarise(total.count = sum (fatal.count))
# sort them in decreasing order
fatal.new <- fatal.new[order(fatal.new$total.count, decreasing = TRUE),]
fatal.new
```
### <span style="color:red;">`FATALITIES`</span> plot
```{r}
g <- ggplot(data = fatal.new, aes(x = fatal.new$EVTYPE, y = fatal.new$total.count, fill=fatal.new$EVTYPE)) 
g <- g + geom_bar(stat = "identity") + scale_fill_hue(c=45, l=80)
g <- g + geom_text(aes(label = fatal.new$total.count, y = fatal.new$total.count))
g <- g + coord_flip()
g <- g + scale_fill_discrete(name="Event")
g <- g + xlab("Event Type") +  ylab("Fatalities Count") 
g <- g + theme_bw()
g
```

### <span style="color:red;">`INJURIES`</span> Processing

```{r}
# get the event have max count and remove duplicates
injur <- new.storm %>% top_n(n = 1, wt = injur.count) %>% distinct()
# summarise the total count of each event. This will give us unqiue injuries.
injur.new <- injur %>% group_by(EVTYPE) %>% summarise(total.count = sum (injur.count))
# sort them in decreasing order
injur.new <- injur.new[order(injur.new$total.count, decreasing = TRUE),]
injur.new
```
### <span style="color:red;">`INJURIES`</span> plot
```{r}
g <- ggplot(data = injur.new, aes(x = injur.new$EVTYPE, y = injur.new$total.count, fill=injur.new$EVTYPE)) 
g <- g + geom_bar(stat = "identity") + scale_fill_hue(c=45, l=80)
g <- g + geom_text(aes(label = injur.new$total.count, y = injur.new$total.count))
g <- g + coord_flip()
g <- g + scale_fill_discrete(name="Event")
g <- g + xlab("Event Type") +  ylab("Injuries Count") 
g <- g + theme_bw()
g
```
Both plots clearly shows that the most harmful event to the human population health is <span style="color:red;">`TORNADO`</span>.

### <span style="color:red;">`ECONOMIC`</span> Processing

For each event, I will sum individual calculation of property and crop damages.
```{r}
eco.data <- storm.data %>% select(EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)

# change the values as per below
# "","+","-","?" = 1 (for consistency purpose)
# "0","1","2","3","4","5","6","7","8" = 10 (for consistency purpose))
# H|h = 100
# K|k = 1000
# M|m = 1000000
# B|b = 1000000000

eco.data$PROPDMGEXP <- mapvalues(eco.data$PROPDMGEXP, from = c("","+","-","0","1","2","3","4","5","6","7","8","?","B","H","K","M","h","m"), to = c("1","1","1","10","10","10","10","10","10","10","10","10","1","1000000000","100","1000","1000000","100","1000000"))

eco.data$CROPDMGEXP <- mapvalues(eco.data$CROPDMGEXP, from = c("","M","K","m","B","?","0","k","2"), to = c("1","1000000","1000","1000000","1000000000","1","10","1000","10"))

eco.data <- mutate(eco.data, prop.damage = PROPDMG * as.numeric(PROPDMGEXP), crop.damage = CROPDMG * as.numeric(CROPDMGEXP), total.damage = prop.damage + crop.damage)

eco.damage <- aggregate(total.damage ~ EVTYPE, data = eco.data, FUN = sum )
eco.damage <- arrange(df = eco.damage, desc(eco.damage$total.damage))

top.20 <- eco.damage[1:20,]
top.20
```
### <span style="color:red;">`Economic`</span> plot 
```{r}
g <- ggplot(data = top.20, aes(x = top.20$EVTYPE, y = top.20$total.damage/1000000000, fill=top.20$EVTYPE)) 
g <- g + geom_bar(stat = "identity") + scale_fill_hue(c=45, l=80)
g <- g + coord_flip()
g <- g + scale_fill_discrete(name="Event")
g <- g + xlab("Event Type") +  ylab("Total Damage Billion ($)") 
g <- g + theme_bw()
g
```

### Results

1. Events <span style="color:red;">`Tornado, Excessive Heat, Heat`</span> are most harmful w.r.t population health 
2. Events <span style="color:red;">`Flood, Hurricane/Typhoon, Tornado, Storm Surge`</span> cause greatest economic consequences.

*******
